---
output:
  github_document:
    pandoc_args: --webtex
bibliography: datancode.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  out.width = "70%",
  fig.align = "center",
  fig.width = 6,
  fig.asp = .618
  )
options(digits = 2)
```

# Loglinear Models for Three-way tables

> _Note: This is not an `r` package but just a set of `r` functions aiming at one purpose._

## Overview

`R` codes to select the best models fitting three-way tables.

- `dplyr`
- `broom`
- `plyr`

packages would be used. Install these before use this set.

### `_common.R`

1. define `find_xname()`: used in `fitted_val.R`, `goodness_fit.R`
2. modify `broom::tidy.anova()` function

### `model_threeway.R`

function `model_loglin()`

- fit `glm` for given data
- for _every pair of independence model_

### `goodness_fit.R`

function `good_loglin()`

- for fitted `model_loglin`,
- compute **goodness-of-fit**
- implement with `purrr::map()` and `dplyr::bind_rows()`

### `fitted_val.R`

function `fit_loglin()`

- for fitted `model_loglin`,
- compute **fitted values** for _every pair of independence model_
- implement with `purrr::map()` and `plyr::join_all()`

***

Frow now on, this document will show how these functions can be applied.

1. call `tidyverse` and `broom` libraries, and the other functions with `r/_common.R`
2. fit with `model_loglin()`
3. goodness-of-fit with `good_loglin()`
4. choosing the best model based on the goodness-of-fit statistic
5. compute fitted values with `fitted_val`

## Beginning

```{r}
source("r/_common.R")
```

`tidyverse` and `broom` are loaded. Here we should modify `tidy` function of @broom. `tidy.anova()` is not defined for `anova.glm`, so we should add more elements in `renamers` in the function. Using the original `tidy` gives `warning message`.

```
renamers <- c(
    ...
    "Ref.df" = "ref.df",
    "Deviance" = "deviance", # glm object
    "Resid. Df" = "resid.df", # glm object
    "Resid. Dev" = "resid.dev" # glm object
  )
```

In addition, `find_xname()` function would be used later to construct the various independence models.

## Data: Alcohol, cigarette, and marijuana use {-}

Data from @Agresti:2012aa would be used.

> refers to a survey by the Wright State University School of Medicine and the United Health Services in Dayton, Ohio. The survey asked 2276 students in their final year of high school in a nonurban area near Dayton, Ohio, whether they had ever used alcohol, cigarettes, or maijuana.

```{r, message=FALSE}
(substance <-
  read_delim("data/Substance_use.dat", delim = " ") %>% 
  mutate(alcohol = str_trim(alcohol)) %>% 
  mutate_if(is.character, factor) %>% 
  mutate_if(is.factor, fct_rev))
```

Long data format as above is easy to fit loglinear model. Against wide format, i.e. contingency table, try `tidyr::gather()`.

### Change to the long data format

For example, look at the below table.

```{r, message=FALSE}
(aids <- read_table("data/AIDS.dat"))
```

```{r}
aids %>% 
  gather(yes, no, key = symptom, value = count)
```

Finally, we get the long data.

## Fitting GLMs

### in hand

We can write every formula in hand. However, it is annoying.

```{r}
(subs_hierarchy <-
  substance %>% 
  do(
    indep = glm(count ~ alcohol + cigarettes + marijuana, data = ., family = poisson()),
    ac_m = glm(count ~ alcohol + cigarettes + marijuana + alcohol:cigarettes, 
               data = ., family = poisson()),
    amcm = glm(count ~ alcohol + cigarettes + marijuana + alcohol:marijuana + cigarettes:marijuana, 
               data = ., family = poisson()),
    acamcm = glm(count ~ alcohol + cigarettes + marijuana + alcohol:cigarettes + alcohol:marijuana + cigarettes:marijuana, 
                 data = ., family = poisson()),
    acm = glm(count ~ alcohol * cigarettes * marijuana, 
              data = ., family = poisson())
  ))
```

### Defining function

Function can be defined for more general usage. See `r/model_threeway.R`

```{r}
source("r/model_threeway.R")
```

`model_loglin()` fits loglinear model for every pair of independence model.

- `.data`: data
- `yname`: character, response variable
- `glm_fam`: random component of glm, option for `family` of `glm()`. Since we are fitting loglinear model, `poisson()` is set to be default.

It returns `tibble` with list element.

```{r}
(subs_hierarchy <- model_loglin(substance, yname = "count", glm_fam = poisson()))
```

Each list has `glm` object with `[[1]]`.

```{r}
subs_hierarchy$indep[[1]]
```

## Chi-square Goodness-of-fit tests

### Statistic

$$G^2 = 2\sum n_{ijk}\ln\frac{n_{ijk}}{\hat\mu_{ijk}}$$

$$X^2 = \sum\frac{(n_{ijk} - \hat\mu_{ijk})^2}{\hat\mu_{ijk}}$$

with **residual df = the number of cell count - the number of non-redundant parameters**

```{r}
source("r/goodness_fit.R")
```

`good_loglin()` calculates the above statistic for given option `test`. `test = "LRT"` option of `anova.glm()` produces $G^2$. `test = Chisq` gives $X^2$. As noticed, this function _is designed to be_ applied with `purrr::map()` and `dplyr::bind_rows()`.

```{r}
(subs_good <-
  subs_hierarchy %>% 
  map(good_loglin, test = "LRT") %>% 
  bind_rows()) %>% 
  pander::pander()
```

### Choosing the best model

From $G^2$, we compare reduced model to complex model

$$G^2(M_0 \mid M_1) = G^2(M_0) - G^2(M_1) \approx \chi^2\Big(df = df(M_0) - df(M_1)\Big)$$

```{r}
subs_good %>% 
  rename(alternative = model) %>% 
  group_by(resid.df) %>% 
  slice(which.min(resid.dev)) %>% 
  arrange(resid.df) %>% 
  mutate(
    goodness = c(resid.dev[1], -diff(resid.dev)),
    df_good = c(resid.df[1], -diff(resid.df))
  ) %>% 
  mutate(p_value = pchisq(goodness, df = df_good, lower.tail = FALSE)) %>% 
  pander::pander()
```

1. `(AC, AM, CM)` vs saturated model `(ACM)`: cannot reject $M_0$ with p-value `0.5408`, so we choose `(AC, AM, CM)`
2. `(A, CM)` vs `(AC, AM, CM)`: reject $M_0$

Thus, **we use the model `(AC, AM, CM)`**

## Fitted values

```{r}
source("r/fitted_val.R")
```

`fit_loglin()` computes fitted values for each independent model. Use this function with `purrr::map()` and `plyr::join_all()`. In `plyr::join_all()`, every variable name including response is written.

```{r}
subs_hierarchy %>% 
  map(fit_loglin) %>% 
  plyr::join_all(by = c("count", "alcohol", "cigarettes", "marijuana")) %>% 
  pander::pander()
```
